---
title: "MATH 216 Homework 4"
author: "Phil Hoxie"
output: html_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# For data manipulation and visualization
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(knitr))
# For US county and state maps
suppressPackageStartupMessages(library(maps))
# For loading in shapefiles
suppressPackageStartupMessages(library(rgdal))
suppressPackageStartupMessages(library(maptools))
# For interactive maps
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(plotly))
```

## Admistrative:

Please indicate

* Who you collaborated with:Mostly solo, although a bit with Christian and Delany
* Roughly how much time you spent on this HW: 9hrs
* What gave you the most trouble: merging the data to make the Bush map
* Any comments you have: 





## Question 1:

```{r, cache=TRUE, echo=FALSE}
# Load state and county map of US in 2010 from the maps package and convert them
# to data frames so that we can ggplot them.
US_state <- map_data("state") %>% 
  tbl_df()
US_county <- map_data("county") %>% 
  tbl_df()
```

### Choropleth Map of US Voter Behavior in 2000

Download the results of the 2000 election from the [School of Public Affairs at 
American University in DC](http://www.american.edu/spa/ccps/Data-Sets.cfm) and 
create a map involving only the lower 48 states that show voter behavior at a 
**county** level. To keep things simple let's only consider George W. Bush, Al
Gore, and Ralph Nader. Your write-up must include:

1. A [choropleth map](https://en.wikipedia.org/wiki/Choropleth_map) where
    * Counties are filled with red when they tend to favor Bush
    * Counties are filled with white when they tend to be split
    * Counties are filled with blue when they tend to favor Gore and Nader
2. An answer to the following question: which states exhibit the greatest
**within state heterogeneity** in voting? Come up with a mathematical
justification.


### Hints

* Consider `scale_fill_gradient2(name="", low="blue", high="red", mid="white")` 
for the appropriate "mid" point.  See the ggplot2 webpage for [this
command](http://docs.ggplot2.org/0.9.3.1/scale_gradient2.html) for inspiration.
* I believe the county map is from 2010, whereas the election data is from 2000,
as such certain counties will not match up. The number is not that large.
* [Counties in
Virginia](http://en.wikipedia.org/wiki/List_of_counties_in_Virginia)


### Code Hints

This function eliminates all non-alphanumeric characters and spaces and converts
all text to lower case:

```{r}
clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}
clean_text("HeLLO wOrLd.")
```

For the following `ggplot`, see how I didn't define `data` nor any `aes`thetics
in the base `ggplot`, but only within the respective `geom`'s. This is a nice
trick that allows you to have different `data` and different `aes` apply to
different `geom`'s; in this case state-level vs county-level data.

```{r}
ggplot(data=NULL) +
  geom_polygon(data=US_county, aes(x=long, y=lat, group=group, fill=1)) +
  geom_path(data=US_county, aes(x=long, y=lat, group=group), col="black", size=0.05) +
  geom_path(data=US_state, aes(x=long, y=lat, group=group), col="black", size=0.1) +
  coord_map()
```


### Write-Up

```{r, fig.width=12, fig.height=6}
county_results <- read.csv(file = "COUNTY.csv", header = TRUE, sep = ",")
county_results <- county_results %>% 
  mutate(votes = sum(BUSH, GORE, NADER, BUCHANAN, BROWNE, PHILLIPS, WRITEINS, HAGELIN, MCREYNOLDS, HARRIS, DODGE,NOTA, MOOREHEAD, BROWN, VENSON, YOUNGKEIT, LANE, na.rm=TRUE))
  
county_vote <- county_results %>% 
select(CID, STATE, COUNTY, PBUSH, PGORE, PNADER, POTHER, BUSH, GORE, NADER, votes) %>% 
  mutate(bush.share = as.numeric(as.character(PBUSH)), 
         gore.share = as.numeric(as.character(PGORE)), 
         nader.share = as.numeric(as.character(PNADER)), 
         other.share = as.numeric(as.character(POTHER)), 
         bush = as.numeric(BUSH), 
         gore = as.numeric(GORE), 
         nader = as.numeric(NADER),
         votes = as.numeric(votes))

clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}

county_vote_clean <- county_vote %>% 
  mutate(county = clean_text(as.character(COUNTY)), state = as.character(clean_text(STATE))) %>% 
  tbl_df()

counties_clean <- US_county %>% 
  mutate(county = clean_text(subregion), state = clean_text(region)) %>% 
  tbl_df()


county_data <- left_join(counties_clean, county_vote_clean, by = c("county", "state"))

```

```{r, fig.width=12, fig.height=6}
bush.map <- ggplot(data=NULL) +  
  geom_polygon(data=US_county, aes(x=long, y=lat, group=group)) +
  geom_polygon(data=county_data, aes(x=long, y=lat, group=group, fill=bush.share))+  
  geom_path(data=US_county, aes(x=long, y=lat, group=group), col="grey", size=0.00001) +  
  geom_path(data=US_state, aes(x=long, y=lat, group=group), col="black", size=0.0005) +  
  scale_fill_gradient2(midpoint=0.5, low=("Blue"), high=("Red"), mid=("white"))+  
  coord_map() +
  ggtitle("Bush v. Gore 2000 Vote Share by County")
bush.map

```


```{r, fig.width=12, fig.height=6}
state.hetero <- county_vote %>% 
  group_by(STATE) %>% 
  summarise(avg.bush = mean(bush.share), sd.bush = sd(bush.share), total.votes = sum(votes)) %>% 
  mutate(cv = sd.bush/avg.bush)

state.hetero$codes <- as.factor(c("AL","AK","AZ","AR","CA",
               "CO","CT","DE","DC","FL",
               "GA","HI","ID","IL","IN",
               "IA","KS","KY","LA","ME",
               "MD","MA","MI","MN","MS",
               "MO","MT","NE","NV","NY",
               "NH","NJ","NM","NC","ND",
               "OH","OK","OR","PA","RI",
               "SC","SD","TN","TX","UT",
               "VT","VA","WA","WV","WI","WY"))


state.hetero$hover <- with(state.hetero, paste(STATE, '<br>', "Avg", round(avg.bush,4), "Stnd.dev", round(sd.bush,4), "<br>",
                           "votes (10 Thousads)",
                           total.votes/10000))

l <- list(color = toRGB("black"), width = 1)

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

plot_ly(state.hetero, z = cv, text = hover, locations = codes, type = 'choropleth',
        locationmode = 'USA-states', color = cv, colors = 'Reds',
        marker = list(line = l), colorbar = list(title = "County Vote CV")) %>%
  layout(title = 'Heterogenious Voting within States, 2000<br>(Hover for breakdown)', geo = g)
```

The measure of heterogeniety in voting within a state that I used was the coefficient of variation (CV). Within each state, I calculated the coefficient of variation (CV) in Bush vote share among all the counties in that state. The CV is equal to the standard deviation divided by the sample mean. CV = sd/mean. The higher the CV, the more variation within the sample and therefore the greater heterogeneity in voting patterns. 

The traditionaly democratic leaning states (CA, NY, NM, MD) appear darker red on the map thus indicating that they had very high CV's. This is most likely because liberal voters are clustered in big cities, whereas Republicans are more spread out. It appears that the plains state, which trend Republican, have greater homogeneity and therefore a lower CV. 


## Question 2:

In this question, you must make an interactive "Single File" Shiny app that uses
Leaflet. For all 184 census tracts in VT in the 2010 census, present
information on the proportion of the population that is either

* White
* African-American
* Hispanic
* Asian and Pacific Island
* Other, including those of two or more races

Use [Social Explorer](http://www.socialexplorer.com/) to get census data. I did
a demo of this in class. If you don't remember how or are stuck, please speak to
me or get help from your peers. Do not submit copies of the same file.

There should be some mechanism in your Shiny app that allows one the user to toggle
between the different ethnic groups.


### Loading Shapefile Data

Here is some starter code:

```{r}
shapefile_name <- paste(getwd(), "/VT_census_tracts/tl_2015_50_tract.shp", sep="")
VT <- readOGR(shapefile_name, layer = "tl_2015_50_tract", verbose = FALSE)
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1)
```


### Write-Up

Upload your shiny app to the Middlebury Shiny Server (see Lecture 16) and post
the url to the app [here](https://www.nhl.com/), replacing the nhl.com link with 
the link to your app.

Comment on general ethnic demographic trends that's more substantive than just
"Vermont is really white."

```{r, fig.width=12, fig.height=6}
shapefile_name <- paste(getwd(), "/VT_census_tracts/tl_2015_50_tract.shp", sep="")
VT <- readOGR(shapefile_name, layer = "tl_2015_50_tract", verbose = FALSE)

vt.race <- read.csv(file = "vermont.races.2010.csv", header=TRUE, sep = ",")

vt.race <- vt.race %>% 
  mutate(white.1 = SE_T054_002/SE_T054_001, 
         black = SE_T054_003/ SE_T054_001, 
         asian = SE_T054_005/SE_T054_001, 
         two = SE_T054_008/SE_T054_001, 
         hisp = SE_T055_010/SE_T054_001)

VT@data <- data.frame(VT@data, vt.race[match(VT@data$GEOID, vt.race$Geo_FIPS),])


pal1 <- colorQuantile("Blues", VT$white.1, n = 5)
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Blues", VT$white.1, n = 5)(white.1)
  ) %>% 
  addLegend(pal = pal1, title = "White Population (2010)", values = ~white.1, opacity = 1)
```

Vermont is really white. However, the white population is less dense around Burlington, which is in the lower quintile compared to other areas of the state. Keep in mind, that these are relative quintiles that represent the percentage of whites in the population relative to the other census tracts. 

```{r, fig.width=12, fig.height=6}
pal2 <- colorQuantile("Greens", VT$black, n = 5)
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Greens", VT$black, n = 5)(black)
  ) %>% 
  addLegend(pal = pal2, title = "African-American Population (2010)", values = ~black, opacity = 1)
```

African Americans tend to cluster in Burlington and its suburbs. 

```{r, fig.width=12, fig.height=6}
pal3 <- colorQuantile("Reds", VT$asian, n = 5)
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Reds", VT$asian, n = 5)(asian)
  ) %>% 
  addLegend(pal = pal3, title = "Asian Population (2010)", values = ~asian, opacity = 1)
```

Asians appear to be relatively distributed throughout the state, however they don't seem to like the North East Kingdom and the eastern side of the Appalachian Mouz

```{r, fig.width=12, fig.height=6}
pal4 <- colorQuantile("Purples", VT$hisp, n = 5)
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Purples", VT$hisp, n = 5)(hisp)
  ) %>% 
  addLegend(pal = pal4, title = "Hispanic Population (2010)", values = ~hisp, opacity = 1)
```

Hispanics, Like African Americans seem to mostly live in Burlingtion. However, they also seem to populate the western border of the state.

```{r, fig.width=12, fig.height=6}
pal5 <- colorQuantile("Oranges", VT$two, n = 5)
leaflet(VT) %>%
  addTiles() %>% 
  addPolylines(color="black", weight=1) %>% 
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    color = ~colorQuantile("Oranges", VT$two, n = 5)(two)
  ) %>% 
  addLegend(pal = pal5, title = "Multi-Racial Population (2010)", values = ~two, opacity = 1)

```

There seems to be a fairly even distribution of people who identify as more than one race. There is a cluster of multiracial people in Burlingon, but the N.E. Kingdom and many of the areas north and east of Montpillier seem to be multiracial. 

